Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-08-08 Thesis/03-chap3.qmd
fix for emi's comment up to 3.9

@@

Note that if more observations are used for estimating the linear regression, the result of @eq-kl-1 will increase, as the integration will be performed over a higher-dimensional space. For a consistent data generating process, $D$ typically increases logarithmically with the number of observations. This behaviour comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.

@@

\noindent where $C$ is a large enough constant keeping the result positive.

@fig-poly-index and @fig-heter-index display the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity. Each residual plot's MVI is computed using @eq-mvi with $C = 10$. When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans. In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases. Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe. @tbl-mvi provides a summary of the MVI usage and it is applicable to other linear regression models.

@@

```{=html}
<table>
 <thead>
  <tr>
   <th style="text-align:left;"> <span data-qmd="Degree of Model Violations"></span> </th>
   <th style="text-align:left;"> <span data-qmd="Range ($C$ = 10)"></span>  </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> <span data-qmd="Strong"></span>  </td>
   <td style="text-align:left;"> <span data-qmd="$\text{MVI} > 8$"></span>  </td>
  </tr>
  <tr>
   <td style="text-align:left;"> <span data-qmd="Moderate"></span>  </td>
   <td style="text-align:left;"> <span data-qmd="$6 < \text{MVI} < 8$"></span> </td>
  </tr>
  <tr>
   <td style="text-align:left;"> <span data-qmd="Weak"></span>  </td>
   <td style="text-align:left;"> <span data-qmd="$\text{MVI} < 6$"></span>  </td>
  </tr>
</tbody>
</table>

@@

```{r}
#| tbl-cap: "Degree of model violations or the strength of the visual signals according to the Model Violation Index (MVI). The constant $C$ is set to be 10."
#| label: tbl-mvi
data.frame(degree = c("Strong", "Moderate", "Weak"),
           range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
  kableExtra::kable(booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Degree of model violations", "Range ($C$ = 10)"))
```

@@

```{r fig-poly-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of non-linearity violations. The Model violations index (MVI) is displayed atop each residual plot. The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$."}
set.seed(1495)
if (!file.exists(here::here("cached_data/poly_index.rds"))) {
  e_sigma_factor <- 3
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  poly_dgp <- phn_model(j = 3, sigma = 0.1^e_sigma_factor)
  ori_dat <- poly_dgp$gen(300)

@@

dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~e_sigma, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10))
```

```{r fig-heter-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of heteroskedasticity violations. The Model violations index (MVI) is displayed atop each residual plot. The heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$."}
set.seed(1495)
if (!file.exists(here::here("cached_data/heter_index.rds"))) {
  b_factor <- 2
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  heter_dgp <- phn_model(include_z = FALSE, b = 0)
  ori_dat <- poly_dgp$gen(300)

@@

dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~nb, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10))
```

@@

To address this issue while adhering to the principle of visual inference, we can compare the estimated distance $\hat{D}$ to the estimated distances for the null plots in a lineup. Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum estimated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the estimated distance for the $i$-th null plot. This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole observer. The estimated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20. Specifically, for lineups comprising $20t$ plots, where $t$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20t-1$. The $p$-value in this case is given by $\frac{1}{20t} + \frac{1}{20t}\sum_{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function.

Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the estimated distance. Consequently, quantiles of the null distribution can be estimated using the sample quantiles available in statistical software such as R, and these quantiles can be utilized for decision-making purposes. The details of the sample quantile computation can be found in @hyndman1996sample. For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the estimated distance for the true residual plot is significantly different from the estimated distance for null plots with a 95% significance level. Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100. However, if the null distribution exhibits a long tail, a larger number of null plots may be required. Alternatively, a $p$-value can be used to represent the probability of observing an event equally or more extreme than the given event under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.

If precision in sample quantiles is not the main priority, using a pre-calculated table of quantiles is an available option. Such a table offers pre-determined quantiles for a specified number of observations. It is generated by assessing numerous null plots derived from various simulated regression models and averaging them. Essentially, this shifts the computational burden from the user to the developer.

@@

Bootstrap is often employed in linear regression when conducting inference for estimated parameters. It is typically done by sampling individual cases with replacement and refitting the regression model. If the observed data accurately reflects the true distribution of the population, the bootstrapped estimates can be used to measure the variability of the parameter estimate without making strong distributional assumptions about the data generating process.

@@

While observational data is frequently employed in training models for real-world applications, the data generating process of observational data often remains unknown, making computation for our target variable $D$ unattainable. Consequently, the computer vision models developed in this study were trained using synthetic data, including 80000 training images and 8000 test images. This approach provided us with precise label annotations. Additionally, it ensured a large and diverse training dataset, as we had control over the data generating process, and the simulation of the training data was relatively cost-effective.

@@

$$
\begin{aligned}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \sqrt{\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^2},
\end{aligned}
$$ {#eq-data-sim}

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{z}$, $\boldsymbol{w}$, $\boldsymbol{k}$ and $\boldsymbol{\varepsilon}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$ are two independent predictors, $\text{He}_j(.)$ is the $j$th-order probabilist's Hermite polynomials [@hermite1864nouveau], the $\sqrt{(.)}$ and $(.)^2$ operators are element-wise operators, $\odot$ is the Hadamard product, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as

$$g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{min}\boldsymbol{1}_n}{x_{max} - x_{min}} - k\boldsymbol{1}_n,~for~k > 0,$$
\noindent where $x_{min} = \underset{i \in \{ 1,...,n\}}{min} x_i$, $x_{max} = \underset{i \in \{ 1,...,n\}}{max} x_i$ and $x_i$ is the $i$-th entry of $\boldsymbol{x}$.

@@

Additionally, Scaling factor $\boldsymbol{k}$ directly affects the error distribution and it is correlated with $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$. If $b \neq 0$ and $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)$, the constant variance assumption will be violated. Parameter $a$ is a shape parameter controlling the location of the smallest variance in a residual plot as shown in @fig-different-a.

@@

Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$. In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and lognormal distributions, as presented in @fig-different-e. Each distribution imparts unique characteristics to the residuals. The discrete error distribution introduces discreteness in residuals, while the lognormal distribution typically yields outliers. Uniform error distribution may result in residuals filling the entire space of the residual plot. All of these distributions exhibit visual distinctions from the normal error distribution.

@@

In real-world analysis, it's not uncommon to encounter instances where multiple model violations coexist. In such cases, the residual plots often exhibit a mixed pattern of visual anomalies corresponding to different types of model violations. @fig-different-j-heter and @fig-different-e-heter show the visual patterns of models with multiple model violations.

### Scagnostics

For each generated residual plot, we computed four scagnostics – "Monotonic", "Sparse", "Splines", and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr]. These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model. While other scagnostics provide valuable insights, they come with high computational costs and are not suitable for quick inference.

@@

With 80000 training images prepared, each bucket accommodated a maximum of $80000 \div 50 = 1600$ training samples. The simulator iteratively sampled parameter values from the parameter space, generated residuals and fitted values using the data generation process, computed the distance, and checked if the sample fitted within the corresponding bucket. This process continued until all buckets were filled.

Similarly, we adopted the same methodology to prepare 8000 test images for performance evaluation and model diagnostics.

@@

The model weights $\boldsymbol{\theta}$ were randomly initialized and they were optimized by the Adam optimizer with the mean square error loss function

$$\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{argmin}\frac{1}{n_{train}}\sum_{i=1}^{n_{train}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,$$

\noindent where $n_{train}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the additional information about the $i$-th residual plot including four scagnostics and the number of observations.

@@

The "number of base filters" determines the number of filters for the first 2D convolutional layer. In the VGG16 architecture, the number of filters for the 2D convolutional layer in a block is typically twice the number in the previous block, except for the last block, which maintains the same number of convolution filters as the previous one. This hyperparameter aids in controlling the complexity of the computer vision model. Higher numbers of base filters result in more trainable parameters. Likewise, the "number of units for the fully-connected layer" determines the complexity of the final prediction block. Increasing the number of units enhances model complexity, resulting in more trainable parameters.

The dropout rate and batch normalization are flexible hyperparameters that work in conjunction with other parameters to facilitate smooth training. A higher dropout rate is necessary when the model tends to overfit the training dataset by learning too much noise. Conversely, a lower dropout rate is preferred when the model is complex and challenging to converge. Batch normalization, on the other hand, addresses the internal covariate shift problem arising from the randomness in weight initialization and input data. It helps stabilize and accelerate the training process by normalizing the activations of each layer.

Additionally, incorporating additional inputs such as scagnostics and the number of observations can potentially enhance prediction accuracy. Therefore, we allowed the tuner to determine whether these inputs were necessary for optimal model performance.

@@

Our model was trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras]. During training, 80% of the training data was utilized for actual training, while the remaining 20% was used as validation data. The Bayesian optimization tuner conducted 100 trials to identify the best hyperparameter values based on validation root mean square error. The tuner then restored the best epoch of the best model from the trials. Additionally, we applied early stopping, terminating the training process if the validation root mean square error fails to improve for 50 epochs. The maximum allowed epochs was set at 2000, although no models reached this threshold.

```{r}
#| label: tbl-hyperparameter
#| tbl-cap: "Name of hyperparameters and their correspoding domain for the computer vision model."
data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
           Domain = c("{4, 8, 16, 32, 64}", "[0.1, 0.6]", "{false, true}", "{max, average}", "{false, true}", "{128, 256, 512, 1024, 2048}", "{false, true}", "[0.1, 0.6]", "[1e-8, 1e-1]")) %>%
  kableExtra::kable(booktabs = TRUE)
```

@@

All optimized models chose to retain the additional inputs, contributing to the reduction of validation error. The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex. The optimized learning rates were smaller than the default value recommended by Keras, which is 0.001.

>>>>

### Auxiliary Information with Scagnostics

For each residual plot used as an input image, we computed four scagnostics – "Monotonic", "Sparse", "Splines", and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr]. These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model. While other scagnostics provide valuable insights, they come with high computational costs and are not suitable for quick inference.

@@

Note that if more observations are used for estimating the linear regression, the result of @eq-kl-1 will increase, as the integration will be performed over a higher-dimensional space. For a given data generating process, $D$ typically increases logarithmically with the number of observations. This behaviour comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.

@@

\noindent where $C$ is a large enough constant keeping the result positive and the term $-\log(n)$ is used to offset the increase in $D$ due to sample size.

@fig-poly-heter-index displays the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity. Each residual plot's MVI is computed using @eq-mvi with $C = 10$. When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans. In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases. Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe. @tbl-mvi provides a summary of the MVI usage and it is applicable to other linear regression models.

@@

```{=html}
<table>
 <thead>
  <tr>
   <th style="text-align:left;"> <span data-qmd="Degree of Model Violations"></span> </th>
   <th style="text-align:center;"><span data-qmd="Range ($C$ = 10)"></span>  </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> <span data-qmd="Strong"></span>  </td>
   <td style="text-align:center;"> <span data-qmd="$\text{MVI} > 8$"></span>  </td>
  </tr>
  <tr>
   <td style="text-align:left;"> <span data-qmd="Moderate"></span>  </td>
   <td style="text-align:center;"> <span data-qmd="$6 < \text{MVI} < 8$"></span> </td>
  </tr>
  <tr>
   <td style="text-align:left;"> <span data-qmd="Weak"></span>  </td>
   <td style="text-align:center;"> <span data-qmd="$\text{MVI} < 6$"></span>  </td>
  </tr>
</tbody>
</table>

@@

```{r}
#| tbl-cap: "Degree of model violations or the strength of the visual signals according to the Model Violation Index (MVI). The constant $C$ is set to be 10."
#| label: tbl-mvi
data.frame(degree = c("Strong", "Moderate", "Weak"),
           range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
  kableExtra::kable(booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Degree of model violations", "Range ($C$ = 10)"),
                    align = "lc") %>%
  kableExtra::kable_styling(latex_options = c("hold_position"))
```

@@

```{r}
set.seed(1495)
if (!file.exists(here::here("cached_data/poly_index.rds"))) {
  e_sigma_factor <- 3
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  poly_dgp <- phn_model(j = 3, sigma = 0.1^e_sigma_factor)
  ori_dat <- poly_dgp$gen(300)

@@

p1 <- dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~e_sigma, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
  ggtitle("(A) Non-linearity")
```

```{r}
set.seed(1495)
if (!file.exists(here::here("cached_data/heter_index.rds"))) {
  b_factor <- 2
  keras_mod <- autovi::get_keras_model("vss_phn_32")
  heter_dgp <- phn_model(include_z = FALSE, b = 0)
  ori_dat <- poly_dgp$gen(300)

@@

p2 <- dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~nb, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10), plot.title = element_text(size = 10)) +
  ggtitle("(B) Heteroskedasticity")
```

```{r fig-poly-heter-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of (A) non-linearity and (B) heteroskedasticity violations. The model violations index (MVI) is displayed atop each residual plot. The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$, while the heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$.", fig.height = 12}
patchwork::wrap_plots(p1, p2, ncol = 1)
```

@@

To address this issue we can adhere to the paradigm of visual inference, by comparing the estimated distance $\hat{D}$ to the estimated distances for the null plots in a lineup. Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum estimated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the estimated distance for the $i$-th null plot. This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole observer. The estimated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.

<!-- For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20. Specifically, for lineups comprising $20t$ plots, where $t$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20t-1$. The $p$-value in this case is given by $\frac{1}{20t} + \frac{1}{20t}\sum_{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function. -->

Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the estimated distance. Consequently, quantiles of the null distribution can be estimated using the sample quantiles, and these quantiles can be utilized for decision-making purposes. The details of the sample quantile computation can be found in @hyndman1996sample. For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the estimated distance for the true residual plot is significantly different from the estimated distance for null plots with a 95% significance level. Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100. However, if the null distribution exhibits a long tail, a larger number of null plots may be required. Alternatively, a $p$-value is the probability of observing a distance equally or greater than $\hat{D}$ under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m} + \frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.

To alleviate computation burden, a lattice of quantiles for $\hat{D}$ under $H_0$ with specified sample sizes can be precomputed. We can then map the $\hat{D}$ and sample size to the closet quantile and sample size in lattice to calculate the corresponding $p$-value. This approach lose precision in $p$-value calculation, however, significantly improves computational efficiency.

@@

Bootstrap is often employed in linear regression when conducting inference for estimated parameters [see @davison1997bootstrap and @efron1994introduction]. It is typically done by sampling individual cases with replacement and refitting the regression model. If the observed data accurately reflects the true distribution of the population, the bootstrapped estimates can be used to measure the variability of the parameter estimate without making strong distributional assumptions about the data generating process.

@@

While observational data is frequently employed in training models for real-world applications, the data generating process of observational data often remains unknown, making computation for our target variable $D$ unattainable. Consequently, the computer vision models developed in this study were trained using synthetic data, including 80,000 training images and 8,000 test images. This approach provided us with precise label annotations. Additionally, it ensured a large and diverse training dataset, as we had control over the data generating process, and the simulation of the training data was relatively cost-effective.

@@

$$
\begin{aligned}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \left[\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^{\circ2}\right]^{\circ1/2},
\end{aligned}
$$ {#eq-data-sim}

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{z}$, $\boldsymbol{w}$, $\boldsymbol{k}$ and $\boldsymbol{\varepsilon}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$ are two independent predictors, $\text{He}_j(.)$ is the $j$th-order probabilist's Hermite polynomials [@hermite1864nouveau], $(.)^{\circ2}$ and $(.)^{\circ1/2}$ are Hadamard square and square root, $\odot$ is the Hadamard product, and $g(\boldsymbol{x}, k)$ is a scaling function to enforce the support of the random vector to be $[-k, k]^n$ defined as

$$A^{\circ\frac12}$$

$$g(\boldsymbol{x}, k) = 2k \cdot \frac{\boldsymbol{x} - x_{\min}\boldsymbol{1}_n}{x_{\max} - x_{\min}} - k\boldsymbol{1}_n,~for~k > 0,$$
\noindent where $x_{\min} = \underset{i \in \{ 1,...,n\}}{\min} x_i$, $x_{\max} = \underset{i \in \{ 1,...,n\}}{\max} x_i$ and $x_i$ is the $i$-th entry of $\boldsymbol{x}$.

@@

Additionally, scaling factor $\boldsymbol{k}$ directly affects the error distribution and it is correlated with $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$. If $b \neq 0$ and $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)$, the constant variance assumption will be violated. Parameter $a$ is a shape parameter controlling the location of the smallest variance in a residual plot as shown in @fig-different-a.

@@

Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$. In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and lognormal distributions, as presented in @fig-different-e. Each distribution imparts unique characteristics in the residual plot. The discrete error distribution introduces discrete clusters in residuals, while the lognormal distribution typically yields outliers. Uniform error distribution may result in residuals filling the entire space of the residual plot. All of these distributions exhibit visual distinctions from the normal error distribution.

@@

In real-world analysis, it is not uncommon to encounter instances where multiple model violations coexist. In such cases, the residual plots often exhibit a mixed pattern of visual anomalies corresponding to different types of model violations. @fig-different-j-heter and @fig-different-e-heter show the visual patterns of models with multiple model violations.

@@

With 80,000 training images prepared, each bucket accommodated a maximum of $80000/ 50 = 1600$ training samples. The simulator iteratively sampled parameter values from the parameter space, generated residuals and fitted values using the data generation process, computed the distance, and checked if the sample fitted within the corresponding bucket. This process continued until all buckets were filled.

Similarly, we adopted the same methodology to prepare 8,000 test images for performance evaluation and model diagnostics.

@@

The model weights $\boldsymbol{\theta}$ were randomly initialized and they were optimized by the Adam optimizer [@kingma2014adam] with the mean square error loss function

$$\hat{\boldsymbol{\theta}} = \underset{\boldsymbol{\theta}}{\text{arg min}}\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(D_i - f_{\boldsymbol{\theta}}(V_i, S_i))^2,$$

\noindent where $n_{\text{train}}$ is the number of training samples, $V_i$ is the $i$-th residual plot and $S_i$ is the additional information about the $i$-th residual plot including four scagnostics and the number of observations.

@@

The number of base filters determines the number of filters for the first 2D convolutional layer. In the VGG16 architecture, the number of filters for the 2D convolutional layer in a block is typically twice the number in the previous block, except for the last block, which maintains the same number of convolution filters as the previous one. This hyperparameter aids in controlling the complexity of the computer vision model. A higher number of base filters results in more trainable parameters. Likewise, the number of units for the fully-connected layer determines the complexity of the final prediction block. Increasing the number of units enhances model complexity, resulting in more trainable parameters.

The dropout rate and batch normalization are flexible hyperparameters that work in conjunction with other parameters to facilitate smooth training. A higher dropout rate is necessary when the model tends to overfit the training dataset by learning too much noise [@srivastava2014dropout]. Conversely, a lower dropout rate is preferred when the model is complex and challenging to converge. Batch normalization, on the other hand, addresses the internal covariate shift problem arising from the randomness in weight initialization and input data [@goodfellow2016deep]. It helps stabilize and accelerate the training process by normalizing the activations of each layer.

Additionally, incorporating additional inputs such as scagnostics and the number of observations can potentially enhance prediction accuracy. Therefore, we allow the tuner to determine whether these inputs were necessary for optimal model performance.

@@

Our model was trained on the MASSIVE M3 high-performance computing platform [@goscinski2014multi], using TensorFlow [@abadi2016tensorflow] and Keras [@chollet2015keras]. During training, 80% of the training data was utilized for actual training, while the remaining 20% was used as validation data. The Bayesian optimization tuner conducted 100 trials to identify the best hyperparameter values based on validation root mean square error. The tuner then restored the best epoch of the best model from the trials. Additionally, we applied early stopping, terminating the training process if the validation root mean square error fails to improve for 50 epochs. The maximum allowed epochs was set at 2,000, although no models reached this threshold.

```{r}
#| label: tbl-hyperparameter
#| tbl-cap: "Name of hyperparameters and their correspoding domain for the computer vision model."
data.frame(`Hyperparameter` = c("Number of base filters", "Dropout rate for convolutional blocks", "Batch normalization for convolutional blocks", "Type of global pooling", "Ignore additional inputs", "Number of units for the fully-connected layer", "Batch normalization for the fully-connected layer", "Dropout rate for the fully-connected layer", "Learning rate"),
           Domain = c("{4, 8, 16, 32, 64}", "[0.1, 0.6]", "{false, true}", "{max, average}", "{false, true}", "{128, 256, 512, 1024, 2048}", "{false, true}", "[0.1, 0.6]", "[$10^{-8}$, $10^{-1}$]")) %>%
  kableExtra::kable(booktabs = TRUE)
```

@@

All optimized models chose to retain the additional inputs, contributing to the reduction of validation error. The number of units required for the fully-connected layer was $256$, a relatively modest number compared to the VGG16 classifier, suggesting that the problem at hand was less complex. The optimized learning rates were higher for models with higher resolution input, likely because models with more parameters are more prone to getting stuck in local minima, requiring a higher learning rate.
